Estimating the phylogenetic age of a fossil

### 2015-08-11 fossilDating.tex

Although there has been much controversy surrounding apparent discrepancies between palaeontological data and molecular phylogenetic inferences it is clear that fossil and molecular data both produce broadly concordant views of evolutionary history\footnote{For example see \url{http://www.actionbioscience.org/evolution/benton.html}}. Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions. In particular, although models of molecular evolution have continually been refined over many decades there has been little change in the modelling of fossil data until very recently. One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record in fact demonstrated concordance between inferences from  molecular data and paleontological data \autocite{tavare2002using}.

@@

The phylogenetic estimation of the age of a taxon based on its molecular sequence has been previously described \autocite{drummond2002computational,shapiro2011bayesian} and applied to both ancient subfossil remains and rapidly evolving viral taxa. For example, this technique has been successfully employed to estimate the age of human subfossil remains based on an ancient mitochondrial genome sequence \autocite{meyer2014mitochondrial}. The same technique has also been used to estimate the age of viral samples based on molecular sequence data \autocite{gray2013evolutionary}.

@@

\textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach to the phylogenetic estimation of time-trees that employed both fossil and molecular sequence data as equal partners under the rule of probability. We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding palaeontological age ranges based on radiometrically-dated geological strata. The model of time-tree phylogeny employed is the so-called fossilized birth-death process \autocite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \autocite{Gavr2013}.

A previous study described how a set of fossils with discrete morphological characters could be used to estimate a phylogeny. Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. This allows for their ages to be estimated solely based on their morphological characters and the phylogenetic affinities of those characters to other fossils with known ages in the phylogeny. We refer to this estimate of a fossil's age as its {\em phylogenetic age}. In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) how does the amount of morphological data available for a new fossil and the number of related reference fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age. These three questions are investigated in the context of a morphological data set of of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).

@@

\begin{figure}
\includegraphics[width=5in]{run8_3/8_phyloAgeVsGeoAge.pdf}
\caption{\label{fig:phyloAgeVsGeoAge}
The Bayesian phylogenetic age estimate (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates. The palaeontological age estimates are represented by the mid-point of the range and the upper and lower limits. The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. Blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. If the horizontal line doesn't cross $x=y$, then the median phylogenetic estimate is not contained in the palaeontological age range. The three labelled fossils show significant inconsistency.}
\end{figure}

\begin{figure}
%\includegraphics[width=5in]{run8_3/8_precisionVsKnownCharacters.pdf}
\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of its phylogenetic age (i.e. the precisions is 1/variance in the marginal posterior distribution of the age).}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{run8_3/fossilDatingDensities_younger.pdf}
\caption{Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr. Red boxes are the superimposed age ranges derived from geological data.}
\end{figure}

@@

% latex table generated in R 3.0.3 by xtable 1.7-4 package
% Tue Jul 28 17:38:04 2015
\begin{table}[ht]
\centering
\begin{tabular}{rlrrrrrr}
  \hline
 & Name & post & Error & Rel Err & Age & Lower & Upper \\ 
  \hline
1 & Anthropornis grandis & 0.932 & 5.05 & 0.12 & 38.22 & 33.17 & 43.42 \\ 
  2 & Anthropornis nordenskjoeldi & 0.903 & 5.01 & 0.12 & 38.30 & 32.04 & 44.33 \\ 
  3 & Archaeospheniscus lopdelli & 0.427 & 1.95 & 0.07 & 26.03 & 21.74 & 31.42 \\ 
  4 & Archaeospheniscus lowei & 0.527 & 0.44 & 0.02 & 27.67 & 21.76 & 34.22 \\ 
  5 & Burnside Palaeudyptes & 0.476 & 1.21 & 0.03 & 36.03 & 33.16 & 39.30 \\ 
  6 & Delphinornis arctowskii & 0.477 & 2.01 & 0.05 & 35.87 & 21.34 & 49.43 \\ 
  7 & Delphinornis gracilis & 0.213 & 6.60 & 0.18 & 44.14 & 35.84 & 51.98 \\ 
  8 & Delphinornis larseni & 0.955 & 1.67 & 0.04 & 40.72 & 33.48 & 51.07 \\ 
  9 & Delphinornis wimani & 0.193 & 15.68 & 0.36 & 26.43 & 17.51 & 41.30 \\ 
  10 & Duntroonornis parvus & 0.771 & 0.12 & 0.00 & 26.00 & 16.07 & 32.26 \\ 
  11 & Eretiscus tonnii & 0.476 & 2.78 & 0.15 & 15.89 & 10.36 & 20.30 \\ 
  12 & Icadyptes salasi & 0.198 & 1.77 & 0.05 & 34.70 & 31.48 & 37.38 \\ 
  13 & Inkayacu paracasensis & 0.387 & 0.79 & 0.02 & 35.84 & 32.49 & 38.48 \\ 
  14 & Kairuku grebneffi & 0.614 & 1.39 & 0.05 & 29.36 & 26.11 & 32.74 \\ 
  15 & Kairuku waitaki & 0.609 & 1.49 & 0.05 & 29.21 & 25.11 & 33.98 \\ 
  16 & Madrynornis mirandus & 0.017 & 3.65 & 0.36 & 6.58 & 2.00 & 9.58 \\ 
  17 & Marambiornis exilis & 0.643 & 1.87 & 0.05 & 39.48 & 33.03 & 46.07 \\ 
  18 & Marplesornis novaezealandiae & 0.377 & 5.55 & 0.52 & 16.54 & 12.41 & 19.85 \\ 
  19 & Mesetaornis polaris & 0.716 & 1.12 & 0.03 & 38.71 & 30.61 & 44.81 \\ 
  20 & Pachydyptes ponderosus & 0.364 & 0.46 & 0.01 & 35.88 & 32.35 & 38.79 \\ 
  21 & Palaeeudyptes antarcticus & 0.148 & 4.49 & 0.14 & 37.36 & 30.67 & 42.56 \\ 
  22 & Palaeeudyptes gunnari & 0.873 & 7.10 & 0.16 & 37.20 & 32.44 & 41.79 \\ 
  23 & Palaeeudyptes klekowskii & 0.857 & 6.22 & 0.14 & 37.42 & 32.46 & 41.78 \\ 
  24 & Palaeospheniscus bergi & 0.932 & 2.47 & 0.16 & 17.77 & 14.11 & 21.95 \\ 
  25 & Palaeospheniscus biloculata & 0.784 & 0.95 & 0.05 & 17.58 & 12.20 & 21.61 \\ 
  26 & Palaeospheniscus patagonicus & 0.931 & 0.18 & 0.01 & 18.53 & 15.61 & 20.66 \\ 
  27 & Paraptenodytes antarcticus & 0.003 & 7.63 & 0.35 & 29.72 & 26.34 & 33.67 \\ 
  28 & Perudyptes devriesi & 0.126 & 5.69 & 0.14 & 45.63 & 39.72 & 53.69 \\ 
  29 & Platydyptes marplesi & 0.768 & 2.17 & 0.08 & 24.43 & 20.72 & 27.79 \\ 
  30 & Platydyptes novaezealandiae & 0.515 & 0.83 & 0.03 & 23.92 & 19.11 & 28.55 \\ 
  31 & Pygoscelis grandis & 0.756 & 0.43 & 0.08 & 5.61 & 1.38 & 10.72 \\ 
  32 & Spheniscus megaramphus & 0.535 & 1.60 & 0.20 & 6.54 & 3.28 & 9.11 \\ 
  33 & Spheniscus muizoni & 0.003 & 4.09 & 0.45 & 4.97 & 2.48 & 7.81 \\ 
  34 & Spheniscus urbinai & 0.882 & 0.01 & 0.00 & 7.82 & 4.99 & 9.95 \\ 
  35 & Waimanu manneringi & 0.068 & 3.31 & 0.05 & 57.91 & 53.32 & 62.91 \\ 
  36 & Waimanu tuatahi & 0.408 & 2.27 & 0.04 & 60.67 & 55.09 & 66.38 \\ 
   \hline
\end{tabular}
\caption{\label{fossilTable}Table of 36 fossil penguins. $post$ is the posterior probability that the phylogenetic age is within the paleaontological age range. Error is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\em Age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\% HPD credible interval.}
\end{table}

>>>>

Although there has been much controversy surrounding apparent discrepancies between palaeontological data and molecular phylogenetic inferences it is clear that fossil and molecular data both produce broadly concordant views of evolutionary history\footnote{For example see \url{http://www.actionbioscience.org/evolution/benton.html}}. Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions. In particular, although models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu} there has been little change in the modelling of fossil data until very recently. One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record in fact demonstrated concordance between inferences from  molecular data and paleontological data \autocite{tavare2002using}.

@@

The phylogenetic estimation of the age of a taxon based on its molecular sequence has been previously described \autocite{drummond2002computational,shapiro2011bayesian} and applied to both ancient subfossil remains and rapidly evolving viral taxa. For example, this technique has been successfully employed to estimate the age of human subfossil remains based on an ancient mitochondrial genome sequence \autocite{meyer2014mitochondrial}. The same technique has also been used to estimate the age of viral samples based on molecular sequence data \autocite[e.g.][]{gray2013evolutionary}.

@@

\textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach to the phylogenetic estimation of time-trees that employed both fossil and molecular sequence data as equal partners under the rule of probability. We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding palaeontological age range based on radiometrically-dated geological strata. The model of time-tree phylogeny employed is the so-called fossilized birth-death process \autocite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \autocite{Gavr2013}.

A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree or chrono-phylogeny. Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. This allows for the age of some of the fossils to be estimated solely based on their morphological characters and the phylogenetic affinities of their morphology to other fossils with known ages in the time-tree. We refer to this estimate of a fossil's age as its {\em phylogenetic age}. In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age. These three questions are investigated in the context of a morphological data set of of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).

@@

\begin{figure}
\includegraphics[width=5in]{run8_5/8_phyloAgeVsGeoAge.pdf}
\caption{\label{fig:phyloAgeVsGeoAge}
The Bayesian phylogenetic age estimate (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates. The palaeontological age estimates are represented by the mid-point of the range and the upper and lower limits. The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. Blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. If the horizontal line doesn't cross $x=y$, then the median phylogenetic estimate is not contained in the palaeontological age range. The three labelled fossils show significant inconsistency.}
\end{figure}

\begin{figure}
\includegraphics[width=5in]{run8_5/8_precisionVsKnownCharacters.pdf}
\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of its phylogenetic age (i.e. the precisions is 1/variance in the marginal posterior distribution of the age).}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{run8_5/fossilDatingDensities_younger.pdf}
\caption{Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr. Red boxes are the superimposed age ranges derived from geological data.}
\end{figure}

@@

\input{run8_5/8_summaryTable.tex}
